---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Build Status](https://travis-ci.org/mpadge/spatialcluster.svg)](https://travis-ci.org/mpadge/spatialcluster) 
[![Project Status: Concept - Minimal or no implementation has been done yet.](http://www.repostatus.org/badges/0.1.0/concept.svg)](http://www.repostatus.org/#concept)
[![codecov](https://codecov.io/gh/mpadge/spatialcluster/branch/master/graph/badge.svg)](https://codecov.io/gh/mpadge/spatialcluster)

# spatialcluster

R port of redcap spatially-constrained clustering, as described in [D. Guo's
2008 paper, "Regionalization with dynamically constrained agglomerative
clutering and
partitioning."](https://www.tandfonline.com/doi/abs/10.1080/13658810701674970)
(pdf available
[here](https://pdfs.semanticscholar.org/ead1/7df8aaa1aed0e433b3ae1ec1ec5c7e785b2b.pdf)).
''Spatially-constrained'' implies that the points which are to be clustered
have defined spatial coordinates, while the clustering itself is based on
additional data, generally in the form of a distance or similarity matrix
quantifying relationships between those points.


## Installation

You can install spatialcluster from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("mpadge/spatialcluster")
```

## Usage

The single function `scl_cluster()` requires three arguments:

1. A rectangular matrix of coordinates of points to be clustered (`n` rows; at
   least 2 columns);
2. An `n`-by-`n` square matrix quantifying relationships between those points;
3. A single value (`ncl`) specifying the desired number of clusters.

Usage can be demonstrated with some simple fake data:
```{r}
n <- 20
xy <- matrix (runif (2 * n), ncol = 2)
dmat <- matrix (runif (n ^ 2), ncol = n)
```
The load the package and call the function:
```{r load, echo = FALSE, message = FALSE}
devtools::load_all (".", export_all = FALSE)
```
```{r fake-run, eval = FALSE}
library (spatialcluster)
scl <- scl_cluster (xy, dmat, ncl = 4)
plot (scl)
```
```{r real-run, echo = FALSE}
scl <- scl_cluster (xy, dmat, ncl = 4)
plot (scl)
```
