---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Build Status](https://travis-ci.org/mpadge/spatialcluster.svg)](https://travis-ci.org/mpadge/spatialcluster) 
[![Project Status: WIP](http://www.repostatus.org/badges/latest/wip.svg)](http://www.repostatus.org/#wip)
[![codecov](https://codecov.io/gh/mpadge/spatialcluster/branch/master/graph/badge.svg)](https://codecov.io/gh/mpadge/spatialcluster)

# spatialcluster

Clustering algorithms are borne of historical computational necessity.
Approximations were sought to boost efficiency and thereby enable clusters to be
computed at all. This package includes an algorithm to *exactly* compute
spatially constrained clusters. The algorithm is as efficient as can be, but is
of course slower than approximate algorithms. It is nevertheless exact.

''Spatially-constrained'' means that the data from which clusters are to be
formed also map on to spatial reference points, and the constraint is that
clusters must be spatially contiguous. The actual clustering data are generally
in the form of a distance or similarity matrix quantifying relationships between
some collection of objects or points, as with most clustering procedures, while
the spatial data are simply coordinates of those objects or points.

For comparison, this package also includes the REDCAP collection of efficient
yet approximate algorithms described in [D. Guo's 2008 paper, "Regionalization
with dynamically constrained agglomerative clustering and
partitioning."](https://www.tandfonline.com/doi/abs/10.1080/13658810701674970)
(pdf available
[here](https://pdfs.semanticscholar.org/ead1/7df8aaa1aed0e433b3ae1ec1ec5c7e785b2b.pdf)).


## Installation

You can install spatialcluster from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("mpadge/spatialcluster")
```

## Usage

The single function `scl_cluster()` requires three arguments:

1. A rectangular matrix of coordinates of points to be clustered (`n` rows; at
   least 2 columns);
2. An `n`-by-`n` square matrix quantifying relationships between those points;
3. A single value (`ncl`) specifying the desired number of clusters.

Usage can be demonstrated with some simple fake data:
```{r}
n <- 100
xy <- matrix (runif (2 * n), ncol = 2)
dmat <- matrix (runif (n ^ 2), ncol = n)
```
The load the package and call the function:
```{r load, echo = FALSE, message = FALSE}
devtools::load_all (".", export_all = FALSE)
```
```{r fake-run, eval = FALSE}
library (spatialcluster)
scl <- scl_cluster (xy, dmat, ncl = 8, linkage = "single")
plot (scl)
```
```{r plot-single, echo = FALSE}
scl <- scl_cluster (xy, dmat, ncl = 8, linkage = "single")
plot (scl, convex = )
```
```{r plot-average}
scl <- scl_cluster (xy, dmat, ncl = 8, linkage = "average")
plot (scl)
```
```{r plot-complete}
scl <- scl_cluster (xy, dmat, ncl = 8, linkage = "complete")
plot (scl)
```

This example illustrates the universal danger in all clustering algorithms: they
can not fail to produce results, even when the data fed to them are definitely
devoid of any information as in this example. Clustering algorithms should only
be applied to reflect a very specific hypothesis for why data should be
clustered in the first place; spatial clustering algorithms should only be
applied to reflect two very specific hypothesis for (i) why data should be
clustered at all, and  (ii) why those clusters should manifest a spatial
pattern.
